\texttt{ggmatplot}: An R package for data visualization on wide-format data

### 2022-05-05 Paper/paper.Rmd
Adding edits and comments to JOSS ms

The layered grammar of graphics [@Wickham2010-kt], implemented as the `ggplot2` package [@Wickham2016] in the statistical language R [@rstats], is a powerful and popular tool to create versatile statistical graphics. This graphical system, however, requires input data to be organised in a manner that a data column is mapped to an aesthetic element (e.g. x-coordinate, y-coordinate, color, size), which create friction in constructing plots with an aesthetic element that span multiple columns in the original data by requiring users to re-organise the data.

The `ggmatplot`, built upon `ggplot2`, is an R-package that allows quick plotting across the columns of matrices or data with the result returned as a `ggplot` object. The package is inspired by the function `matplot()` in the core R `graphics` system, thus `ggmatplot` can be considered as a `ggplot` version of `matplot` with the benefits of customising the plots as any other `ggplot` objects via `ggplot2` functions.

@@

Input data to construct plots with `ggplot2` require data to be organised in a manner that maps data columns to aesthetic elements. This generally works well where data is tidied in a rectangular form, referred to as "tidy data" [@Wickham2014-gy], where each row represents an observational unit, each column represents a variable, and each cell represents a value. In some cases, what constitutes a variable (or observational unit), hence a column (or row), in a tidy data can be dependent upon interpretation or downstream interest (e.g. Tables \ref{tab:tab1} and \ref{tab:tab2} can be both considered as tidy data), but a clear violation of tidy data principles is when the column names contain data values, e.g. Table \ref{tab:tab3} contain the name of the species across a number of column names.

@@

The organisation of the data is largely dependent on the downstream analysis and there is no one correct way to do this. Some forms of multivariate data, e.g. Table \ref{tab:tab3}, are prevalent in the field because it aligns as an input data for a modelling software and/or the format is more convenient for input or view of the data in spreadsheet format. However, this format is not consistent with the required format for `ggplot2`, and consequently, plotting with `ggplot2` interrupts the workflow of a user that is trying to quickly visualise these types of data. The `ggmatplot` R-package provides a solution to this common friction in producing plots with `ggplot2`.

@@

In this section we demonstrate the use of the `ggmatplot` package and contrast the specification with `ggplot2` after data wrangling using `dplyr` and `tidyr` [@Wickham2019] using the example data in Tables \ref{tab:tab1} and \ref{tab:tab3}, which are stored in the objects `wide_df` and `SnowGR`, respectively.

@@

The code below constructs a line plot (superimposed with a point) of the various types (food, service and ambience) of rating, contained in columns 2 to 4 of `wide_df`, against the overall restaurant rating in column 5 of `wide_df` as shown in Figure 1.

@@

In contrast to the above, the data must be wrangled to a long form first before plotting in the code below to obtain the same figure as Figure 1. This adds a small, but noticable, friction to the workflow for the user that is looking to quickly explore their data.

@@

The example code draws the boxplot of each column of amount of snowfall across months in the `SnowGR` data as shown in Figure 2. As the resulting object is a `ggplot` object, the user can leverage the `ggplot` functions to modify the output (e.g. removal of the legend).

@@

The equivalent code for the above to produce Figure 2 without using `ggmatplot` is given below. There is again a slight friction in putting the data in the right format before plotting. The original wide data format like those shown in Table \ref{tab:tab3} is common so an analyst who has to repeat these tasks can benefit from a quick approach.

@@

@@

The `ggmatplot` R-package provides a solution to a common friction to quickly plotting multivariate data where the primary interest is mapping the column names as an aesthetic element. The solution provided however is a recipe-driven approach where the user can only produce plot types as many there are included in the `plot_type` option. Future development of the package could benefit from using a grammar approach, like in @Wilkinson2005-oz and @Wickham2010-kt, where plot types can be extensible.

@@

FKCH was supported by ARC DECRA XXX.

>>>>

The layered grammar of graphics [@Wickham2010-kt], implemented as the `ggplot2` package [@Wickham2016] in the statistical language R [@rstats], is a powerful and popular tool to create versatile statistical graphics. However, this graphical system requires input data to be organised in a manner that a data column is mapped to an aesthetic element (e.g. x-coordinate, y-coordinate, color, size), which creates friction in constructing plots with an aesthetic element that span multiple columns in the original data by requiring users to re-organise the data.  
<!-- Should we use the words long and wide format data frames here, or is that wrong? I only ask because I (maybe others?) think about ggplot requiring long format data, but we want to wide formatting? -->

The `ggmatplot`, built upon `ggplot2`, is an R-package that allows quick plotting across the columns of matrices or data with the result returned as a `ggplot` object. The package is inspired by the function `matplot()` in the core R `graphics` system -- as such, `ggmatplot` may be considered as a `ggplot` version of `matplot` with the benefits of customising the plots as any other `ggplot` objects via `ggplot2` functions, as well as offering several other plotting types that are not immediately available from `matplot` directly, such as comparative violin plots.

@@

Input data to construct plots with `ggplot2` require data to be organised in a manner that maps data columns to aesthetic elements. This generally works well where data is tidied in a long rectangular form, often referred to as "tidy data" [@Wickham2014-gy], where each row represents an observational unit, each column represents a variable, and each cell represents a value. In some cases, what constitutes a variable (or observational unit), and hence a column (or row), in tidy data can be dependent upon interpretation or downstream interest (e.g. Tables \ref{tab:tab1} and \ref{tab:tab2} can be both considered as tidy data), but a clear violation of tidy data principles is when the column names contain data values, e.g. *Table \ref{tab:tab3} contain the name of the species across a number of column names.* THIS IS WRONG? THERE ARE NO SPECIES HERE?

@@

The organisation of the data is largely dependent on the downstream analysis, and there is no one correct way to do this. Some forms of multivariate data, e.g. Table \ref{tab:tab3}, are prevalent in many scientific fields because it aligns with the input data for a particular modelling software, and/or the format is more convenient for input or view of the data in spreadsheet format (say). Unfortunately, this format is not consistent with the required format for `ggplot2`, and consequently plotting with `ggplot2` interrupts the workflow of a user that is trying to quickly visualise these types of data (as part of their exploratory data analysis, say). The `ggmatplot` R-package seeks to provide a solution to this common friction in producing plots with `ggplot2`.

@@

In this section, we demonstrate the use of the `ggmatplot` package and contrast the specification with `ggplot2` after data wrangling using `dplyr` and `tidyr` [@Wickham2019]. We will use the example data in Tables \ref{tab:tab1} and \ref{tab:tab3}, which are stored in the objects `wide_df` and `SnowGR`, respectively.

@@

The code below constructs a line plot (superimposed with a point) of the various types (food, service and ambience) of ratings, contained in columns 2 to 4 of `wide_df`, against the overall restaurant rating in column 5 of `wide_df` as shown in Figure 1.

@@

In contrast to the above, using `ggplot2` alone, the data must be wrangled to a long form first before plotting, as exemplified in the code below, in order to obtain the same result as Figure 1. This adds a small, but noticeable, friction to the workflow for the practitioner that is looking to promptly explore their data.

@@

The example code draws the boxplot of each column of amount of snowfall across months in the `SnowGR` data as shown in Figure 2. As the resulting object is a `ggplot` object, the user can leverage the `ggplot` functions to modify the output (e.g. removal of the legend). CAN WE `fct_inorder` SO THAT THE MONTHS APPEAR IN ORDER OF THEIR COLUMN NAMES? EMI HAS ALREADY PICKED THIS UP AS AN ISSUE...

@@

The equivalent code for the above to produce Figure 2 without using `ggmatplot` is given below. Again, we observe a slight but non-negligible friction in putting the data in the right format prior to plotting. The original wide data format like those shown in Table \ref{tab:tab3} is common in the environmental sciences among other disciplines, and thus an analyst who has to repeat these tasks can benefit from a quick approach as `ggmatplot` offers.

@@

WOULD IT BENEFIT FROM ALSO SHOWING A PLOT THAT BASE MATPLOT CAN NOT? THE ABOVE CAN BE DONE BY APPLY `boxplot`, AND HENCE WHY I WROTE VIOLIN PLOT AT THE BEGINNING AS IT IS NOT AVAILABLE AT LEAST IN CORE R?

@@

The `ggmatplot` R-package provides a solution to a common friction encountered when wanting to quickly plot multivariate data, where the primary interest is mapping the column names as an aesthetic element. While an excellent start, we also acknowledge that solution provided is a recipe-driven approach, where the user can only produce plot types as many there are included in the `plot_type` option. Future developments of the package could benefit from using a grammar approach, like in @Wilkinson2005-oz and @Wickham2010-kt, where plot types can be extensible.

@@

FKCH was supported by an Australian Research Council Discovery Fellowship DE200100435.

### 2022-05-05 Paper/paper.tex
Adding edits and comments to JOSS ms

% --- Header / Footer ---------------------------------------------------------
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
%\renewcommand{\headrulewidth}{0.50pt}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{\hspace{-0.75cm}\includegraphics[width=5.5cm]{/Library/Frameworks/R.framework/Versions/4.1/Resources/library/rticles/rmarkdown/templates/joss/resources/JOSS-logo.png}}
\fancyhead[C]{}
\fancyhead[R]{}
\renewcommand{\footrulewidth}{0.25pt}

@@

\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={ggmatplot: An R package for data visualization on wide-format data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

The layered grammar of graphics (Wickham, 2010), implemented as the
\texttt{ggplot2} package (Wickham, 2016) in the statistical language R
(R Core Team, 2021), is a powerful and popular tool to create versatile
statistical graphics. This graphical system, however, requires input
data to be organised in a manner that a data column is mapped to an
aesthetic element (e.g.~x-coordinate, y-coordinate, color, size), which
create friction in constructing plots with an aesthetic element that
span multiple columns in the original data by requiring users to
re-organise the data.

The \texttt{ggmatplot}, built upon \texttt{ggplot2}, is an R-package
that allows quick plotting across the columns of matrices or data with
the result returned as a \texttt{ggplot} object. The package is inspired
by the function \texttt{matplot()} in the core R \texttt{graphics}
system, thus \texttt{ggmatplot} can be considered as a \texttt{ggplot}
version of \texttt{matplot} with the benefits of customising the plots
as any other \texttt{ggplot} objects via \texttt{ggplot2} functions.

@@

Input data to construct plots with \texttt{ggplot2} require data to be
organised in a manner that maps data columns to aesthetic elements. This
generally works well where data is tidied in a rectangular form,
referred to as ``tidy data'' (Wickham, 2014), where each row represents
an observational unit, each column represents a variable, and each cell
represents a value. In some cases, what constitutes a variable (or
observational unit), hence a column (or row), in a tidy data can be
dependent upon interpretation or downstream interest (e.g.~Tables
\ref{tab:tab1} and \ref{tab:tab2} can be both considered as tidy data),
but a clear violation of tidy data principles is when the column names
contain data values, e.g.~Table \ref{tab:tab3} contain the name of the
species across a number of column names.

@@

The organisation of the data is largely dependent on the downstream
analysis and there is no one correct way to do this. Some forms of
multivariate data, e.g.~Table \ref{tab:tab3}, are prevalent in the field
because it aligns as an input data for a modelling software and/or the
format is more convenient for input or view of the data in spreadsheet
format. However, this format is not consistent with the required format
for \texttt{ggplot2}, and consequently, plotting with \texttt{ggplot2}
interrupts the workflow of a user that is trying to quickly visualise
these types of data. The \texttt{ggmatplot} R-package provides a
solution to this common friction in producing plots with
\texttt{ggplot2}.

@@

In this section we demonstrate the use of the \texttt{ggmatplot} package
and contrast the specification with \texttt{ggplot2} after data
wrangling using \texttt{dplyr} and \texttt{tidyr} (Wickham et al., 2019)
using the example data in Tables \ref{tab:tab1} and \ref{tab:tab3},
which are stored in the objects \texttt{wide\_df} and \texttt{SnowGR},
respectively.

@@

The code below constructs a line plot (superimposed with a point) of the
various types (food, service and ambience) of rating, contained in
columns 2 to 4 of \texttt{wide\_df}, against the overall restaurant
rating in column 5 of \texttt{wide\_df} as shown in Figure 1.

@@

In contrast to the above, the data must be wrangled to a long form first
before plotting in the code below to obtain the same figure as Figure 1.
This adds a small, but noticable, friction to the workflow for the user
that is looking to quickly explore their data.

@@

The example code draws the boxplot of each column of amount of snowfall
across months in the \texttt{SnowGR} data as shown in Figure 2. As the
resulting object is a \texttt{ggplot} object, the user can leverage the
\texttt{ggplot} functions to modify the output (e.g.~removal of the
legend).

@@

The equivalent code for the above to produce Figure 2 without using
\texttt{ggmatplot} is given below. There is again a slight friction in
putting the data in the right format before plotting. The original wide
data format like those shown in Table \ref{tab:tab3} is common so an
analyst who has to repeat these tasks can benefit from a quick approach.

@@

@@

The \texttt{ggmatplot} R-package provides a solution to a common
friction to quickly plotting multivariate data where the primary
interest is mapping the column names as an aesthetic element. The
solution provided however is a recipe-driven approach where the user can
only produce plot types as many there are included in the
\texttt{plot\_type} option. Future development of the package could
benefit from using a grammar approach, like in Wilkinson (2005) and
Wickham (2010), where plot types can be extensible.

>>>>

% --- Header / Footer ---------------------------------------------------------
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
%\renewcommand{\headrulewidth}{0.50pt}
\renewcommand{\headrulewidth}{0pt}
\fancyhead[L]{\hspace{-0.75cm}\includegraphics[width=5.5cm]{/home/fkch/R/x86_64-pc-linux-gnu-library/3.6/rticles/rmarkdown/templates/joss/resources/JOSS-logo.png}}
\fancyhead[C]{}
\fancyhead[R]{}
\renewcommand{\footrulewidth}{0.25pt}

@@

\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={ggmatplot: An R package for data visualization on wide-format data},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% Pandoc syntax highlighting
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

The layered grammar of graphics (Wickham, 2010), implemented as the
\texttt{ggplot2} package (Wickham, 2016) in the statistical language R
(R Core Team, 2021), is a powerful and popular tool to create versatile
statistical graphics. However, this graphical system requires input data
to be organised in a manner that a data column is mapped to an aesthetic
element (e.g.~x-coordinate, y-coordinate, color, size), which creates
friction in constructing plots with an aesthetic element that span
multiple columns in the original data by requiring users to re-organise
the data.\\

The \texttt{ggmatplot}, built upon \texttt{ggplot2}, is an R-package
that allows quick plotting across the columns of matrices or data with
the result returned as a \texttt{ggplot} object. The package is inspired
by the function \texttt{matplot()} in the core R \texttt{graphics}
system -- as such, \texttt{ggmatplot} may be considered as a
\texttt{ggplot} version of \texttt{matplot} with the benefits of
customising the plots as any other \texttt{ggplot} objects via
\texttt{ggplot2} functions, as well as offering several other plotting
types that are not immediately available from \texttt{matplot} directly,
such as comparative violin plots.

@@

Input data to construct plots with \texttt{ggplot2} require data to be
organised in a manner that maps data columns to aesthetic elements. This
generally works well where data is tidied in a long rectangular form,
often referred to as ``tidy data'' (Wickham, 2014), where each row
represents an observational unit, each column represents a variable, and
each cell represents a value. In some cases, what constitutes a variable
(or observational unit), and hence a column (or row), in tidy data can
be dependent upon interpretation or downstream interest (e.g.~Tables
\ref{tab:tab1} and \ref{tab:tab2} can be both considered as tidy data),
but a clear violation of tidy data principles is when the column names
contain data values, e.g.~\emph{Table \ref{tab:tab3} contain the name of
the species across a number of column names.} THIS IS WRONG? THERE ARE
NO SPECIES HERE?

@@

The organisation of the data is largely dependent on the downstream
analysis, and there is no one correct way to do this. Some forms of
multivariate data, e.g.~Table \ref{tab:tab3}, are prevalent in many
scientific fields because it aligns with the input data for a particular
modelling software, and/or the format is more convenient for input or
view of the data in spreadsheet format (say). Unfortunately, this format
is not consistent with the required format for \texttt{ggplot2}, and
consequently plotting with \texttt{ggplot2} interrupts the workflow of a
user that is trying to quickly visualise these types of data (as part of
their exploratory data analysis, say). The \texttt{ggmatplot} R-package
seeks to provide a solution to this common friction in producing plots
with \texttt{ggplot2}.

@@

In this section, we demonstrate the use of the \texttt{ggmatplot}
package and contrast the specification with \texttt{ggplot2} after data
wrangling using \texttt{dplyr} and \texttt{tidyr} (Wickham et al.,
2019). We will use the example data in Tables \ref{tab:tab1} and
\ref{tab:tab3}, which are stored in the objects \texttt{wide\_df} and
\texttt{SnowGR}, respectively.

@@

The code below constructs a line plot (superimposed with a point) of the
various types (food, service and ambience) of ratings, contained in
columns 2 to 4 of \texttt{wide\_df}, against the overall restaurant
rating in column 5 of \texttt{wide\_df} as shown in Figure 1.

@@

In contrast to the above, using \texttt{ggplot2} alone, the data must be
wrangled to a long form first before plotting, as exemplified in the
code below, in order to obtain the same result as Figure 1. This adds a
small, but noticeable, friction to the workflow for the practitioner
that is looking to promptly explore their data.

@@

The example code draws the boxplot of each column of amount of snowfall
across months in the \texttt{SnowGR} data as shown in Figure 2. As the
resulting object is a \texttt{ggplot} object, the user can leverage the
\texttt{ggplot} functions to modify the output (e.g.~removal of the
legend). CAN WE \texttt{fct\_inorder} SO THAT THE MONTHS APPEAR IN ORDER
OF THEIR COLUMN NAMES? EMI HAS ALREADY PICKED THIS UP AS AN
ISSUE\ldots{}

@@

The equivalent code for the above to produce Figure 2 without using
\texttt{ggmatplot} is given below. Again, we observe a slight but
non-negligible friction in putting the data in the right format prior to
plotting. The original wide data format like those shown in Table
\ref{tab:tab3} is common in the environmental sciences among other
disciplines, and thus an analyst who has to repeat these tasks can
benefit from a quick approach as \texttt{ggmatplot} offers.

@@

WOULD IT BENEFIT FROM ALSO SHOWING A PLOT THAT BASE MATPLOT CAN NOT? THE
ABOVE CAN BE DONE BY APPLY \texttt{boxplot}, AND HENCE WHY I WROTE
VIOLIN PLOT AT THE BEGINNING AS IT IS NOT AVAILABLE AT LEAST IN CORE R?

@@

The \texttt{ggmatplot} R-package provides a solution to a common
friction encountered when wanting to quickly plot multivariate data,
where the primary interest is mapping the column names as an aesthetic
element. While an excellent start, we also acknowledge that solution
provided is a recipe-driven approach, where the user can only produce
plot types as many there are included in the \texttt{plot\_type} option.
Future developments of the package could benefit from using a grammar
approach, like in Wilkinson (2005) and Wickham (2010), where plot types
can be extensible.
